---
title: Uncovering the relationships among functions in a package
author: ~
date: '2018-03-03'
slug: uncovering-the-relationships-among-functions-in-a-package
categories:
  - R
tags:
  - networks
  - data-viz
  - packages
  - DiagrammeR

draft: false
---

I maintain a few internal R packages that we have at work. Recently, and perhaps
foolishly, I decided that I needed to totally refactor one of them to add some more
bells and whistles. A substantial part of this task involved simplifying the 
internal workflow of the package. I needed to be sure not to omit any of the 
previous functionality, but a simple list of all the exported and unexported 
functions was not sufficient.[^1]

[^1]: Also developing this function was a fun way to put off doing the hard work.

So, I decided to map the dependencies between all of the internal functions 
(e.g. which functions called which). I found it useful, so I thought I'd share 
a function I wrote that does just that: visualizes the relationships among functions 
in a package. The functions are available at
[this gist](https://gist.github.com/carlbfrederick/b30d861ea80a27fad4e44623c41e0170).


For the sake of this post, I will use the 
[`merTools`](https://cran.r-project.org/web/packages/merTools/index.html) package 
that I have had the pleasure of working on with a great co-author. The end results are:

```{r, results = 'asis', echo = TRUE, message = FALSE, warning = FALSE}
suppressMessages(
  devtools::source_gist("b30d861ea80a27fad4e44623c41e0170", filename = "packageFunctionMap.R")  
)
fake <- plotFcnDependencies("merTools")
fake %>% select_nodes_by_degree(expressions = "deg > 0") %>% transform_to_subgraph_ws %>% render_graph
```

This diagram shows which of the `merTools` functions call other `merTools` functions. Well,
technically it shows all of the functions that explicitly call other functions. It does 
**not** include dependencies that are based on the output of other functions and/or 
functions that do not call other internal functions. The latter would have degree = 0 in 
social network terms.

Columns to the left are upstream and columns to the right are downstream. For example `REimpact()` 
calls `expectedRank()` and `predictInterval()`. `predictInterval()` then goes on to 
call three more functions, etc.  

The practical value of this exercise comes from the following sorts of insights:
  
  - `stripAttributes()` is called by both `averageObs()` and `randomObs()`, don't forget 
    to include that functionality in whatever function(s) replace those two; and
  - `fastdisp()` is the only function that calls `easyVarCorr()`, perhaps `easyVarCorr()`
    is a good candidate to be rolled into `fastdisp()` if there are no other use cases 
    for this function.

## Stepping through the functions

The first function, `ls_fcns()`, retrieves the names of the functions from a namespace (using 
`lsf.str`) and returns it as a character vector.

```{r ls_fcns, echo=TRUE}
ls_fcns <- function(pkg) {
  fcns <- unclass(lsf.str(envir = asNamespace(pkg), all = TRUE))
  return(as.character(fcns))
}

ls_fcns("merTools")
```

The second function, `fcn_deps()`, cycles through each function to and does a regular expression
search to see which, if any, of the other functions from that namespace are called.
This part could be written more efficiently, I'm sure, but I was starting to feel
bad about procrastinating too much.  Feel free to suggest improvements in the 
comments or via twitter -- [\@carlbfrederick](https://twitter.com/carlbfrederick).

```{r fcn_deps, echo=TRUE}
fcn_deps <- function(pkg) {
  fcns <- ls_fcns(pkg)
  out <- data.frame(Function = as.character(),
                    Dependency_Function = as.character(),
                    Number_Calls = as.integer())
  for (i in fcns) {
    this_fcn <- capture.output(getAnywhere(i))
    for (j in fcns[-grep(i, fcns)]) {
      dep_fcns <- grep(paste(j, "\\(", sep=""), this_fcn)
      if (length(dep_fcns > 0)) {
        out <- rbind(out,
                     data.frame(Function = i,
                                Dependency_Function = j,
                                Number_Calls = length(dep_fcns)))
      }
    }
  }
  return(out)
}

fcn_deps("merTools") %>% as_data_frame
```

As you can see, `fcn_deps()` returns a data.frame with three columns: Function, 
Dependency_Function and Number_Calls. The first two are used to create the diagram.
I don't recall what I originally intended to do with Number_Calls, but it is
still there.

The third function, `plotFcnDependencies()`, produces the
[`DiagrammeR`](https://cran.rstudio.com/web/packages/DiagrammeR/index.html) [@Ianone]  
object that can be plotted. It is definitely "hacky" and will throw some errors 
that I haven't quite ironed out yet, hence the `try()` in there.  

The for loop is used to place the functions horizontally in the diagram above
by calculating the longest of the shortest paths from each function to all the
others. For example, the longest shortest paths between `draw.MerMod()` and 
the rest of the functions is 4. I call this quantity depth in the function.

Using this depth quantity, the rest of the function just creates the node and
edge data.frames and then creates the diagram using the `DiagrammeR` package
... and voila!

```{r, echo=TRUE}
plotFcnDependencies <- function(pkg) {
  fcns <- ls_fcns(pkg)
  depFcn <- fcn_deps(pkg)

  depth <- NULL

  for (i in fcns) {
    try(suppressWarnings(dist <- max(sapply(igraph::shortest_paths(igraph::graph_from_data_frame(depFcn[,1:2]), from = i)$vpath, length))))
    depth <- c(depth, dist)
    dist <- 0
  }

  nodes <- data.frame(nodes = fcns,
                      type = "",
                      label = htmltools::htmlEscape(fcns),
                      depth = depth)

  nodes <- dplyr::arrange(nodes, desc(depth))


  out <- DiagrammeR::create_graph(
    nodes_df = nodes,
    edges_df = DiagrammeR::create_edges(
      from = depFcn$Function,
      to = depFcn$Dependency_Function,
      rel = "leading_to"
    ),
    graph_name = paste(pkg, " (version ", packageVersion(pkg), ") Function Map", sep=""),
    graph_attrs = c("layout = dot", "rankdir = LR"),
    node_attrs = "fontsize = 20"
  )

  return(out)
}
```

## Advice

Many packages have *a lot* of functions. Mapping an entire package often results in an unreadable
network diagram. Go ahead an make a diagram of `dplyr` once and you will see what I mean.

## Ideas for further enhancements

  1. Differentiate exported functions from unexported functions.
  2. Add an argument to focus on certain function(s) to focus the diagram on the area of interest. 

Thanks for reading! Feel free to join the discussion below.

```{r}
devtools::session_info()
```

