---
title: Benchmarking Random Forest Algorithms
author: Carl Frederick
date: '2018-05-30'
slug: benchmarking-random-forest-algorithms
categories:
  - R
tags: []
draft: true
---

## Packages

### [From `caret` documentation](http://topepo.github.io/caret/available-models.html)

  - randomForest
  - ranger
  - extraTrees
  - obligueRF
  - party
  - quantregForest
  - inTrees
  - Rborist
  - RRF
  - wsrf
  
### [From Maching Learning taskview on CRAN](https://cran.r-project.org/view=MachineLearning)

parenthetical for now

  - ipred
  - randomForestSRC
  - LogicForest
  - RLT
  - trtf
  - grf
  
## Choosing some "Training" Data

I will be using the `diamonds` data from the `ggplot2` package as some 
surrogate training data. Since I am only interested in speed/object size I 
don't need any test data. The `diamonds` data are similar in size (number of 
rows, especially) to the data I am using. The next chunk will load those data 
and all the relevant packages used in this post. I will also create a binary 
variable, to mirror my particular use case.

```{r setup, message = FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(caret)
})

data("diamonds", package = "ggplot2")
diamonds$price <- ifelse(diamonds$price > quantile(diamonds$price, probs = .7),
                         "Expensive", "Regular")

trCtrl <- trainControl(method = "repeatedcv", 
                       number = 5, 
                       repeats = 3, 
                       savePredictions = FALSE, 
                       classProbs = TRUE, 
                       summaryFunction = twoClassSummary)
```

```{r rf}
t.rf <- system.time({
  mod1 <- train(price ~ .,
                data = diamonds,
                method = "rf", 
                metric = "ROC", 
                trControl = trCtrl, 
                tuneLength = 5)
})

s.rf <- pryr::object_size(mod1)

```


```{r ranger}
t.ranger <- system.time({
  mod2 <- train(price ~ .,
                data = diamonds,
                method = "ranger", 
                metric = "ROC", 
                trControl = trCtrl, 
                tuneLength = 5)
})

s.ranger <- pryr::object_size(mod2)

```

```{r Rborist}
t.Rborist <- system.time({
  mod3 <- train(price ~ .,
                data = diamonds,
                method = "Rborist", 
                metric = "ROC", 
                trControl = trCtrl, 
                tuneLength = 5)
})

s.Rborist <- pryr::object_size(mod3)

```

